Engineering spin squeezing in a 3D optical lattice with interacting spin-orbit-coupled fermions

### 2019-08-30 soc_squeezing/squeezing_paper/main.tex
some changes before submission

\documentclass[aps,prx,superscriptaddress,twocolumn]{revtex4-2}

@@

%%% text for affiliations and contributions
\newcommand{\JILA}{JILA, National Institute of Standards and Technology and Department of Physics, University of Colorado, Boulder, CO, 80309, USA}
\newcommand{\CTQM}{Center for Theory of Quantum Matter, University of Colorado, Boulder, CO, 80309, USA}
\newcommand{\contrib}{\thanks{Authors P.H.~and M.A.P.~contributed equally to this work.}}

\begin{document}

@@

\author{P.~He} \contrib
\author{M.~A.~Perlin} \contrib
\author{S.~R.~Muleady}
\author{R.~J.~Lewis-Swan}
\affiliation{\JILA}
\affiliation{\CTQM}
\author{R.~B.~Hutson}
\author{J.~Ye}
\affiliation{\JILA}
\author{A.~M.~Rey}
\affiliation{\JILA}
\affiliation{\CTQM}

\begin{abstract}
One of the most important tasks in modern quantum science is to coherently control and entangle many-body systems, and to subsequently use these systems to realize powerful quantum technologies such as quantum-enhanced sensors.
However, many-body entangled states are difficult to prepare and preserve since internal dynamics and external noise rapidly degrade any useful entanglement.
Here, we introduce a protocol that counterintuitively exploits inhomogeneities, a typical source of dephasing in a many-body system, in combination with interactions to generate metrologically useful and robust many-body entangled states.
Motivated by current limitations in state-of-the-art three-dimensional (3D) optical lattice clocks (OLCs) operating at quantum degeneracy, we use local interactions in a Hubbard model with spin-orbit coupling to achieve a spin-locking effect.
In addition to prolonging inter-particle spin coherence, spin-locking transforms the dephasing effect of spin-orbit coupling into a collective spin-squeezing process that can be further enhanced by applying a modulated drive.
Our protocol is fully compatible with state-of-the-art 3D OLC interrogation schemes and may be used to improve their sensitivity, which is currently limited by the intrinsic quantum noise of independent atoms.
We demonstrate that even with realistic experimental imperfections, our protocol may generate $\sim10$--$14$ dB of spin squeezing in $\sim1$ second with $\sim10^2$--$10^4$ atoms.
This capability exemplifies a new paradigm of using driven non-equilibrium systems to overcome current limitations in quantum metrology, allowing OLCs to enter a new regime of enhanced sensing with correlated quantum states.
\end{abstract}

% keywords: ultracold atoms, optical lattices, atomic clocks, metrology, spin squeezing

\maketitle

\section{Introduction}

@@

Optical lattice clocks in particular have seen some of the most impressive developments in recent years, reaching record levels of precision ($\sim 3\times 10^{-19}$)\cite{campbell2017fermidegenerate, marti2018imaging} and accuracy ($\sim 1\times 10^{-18}$)\cite{bloom2014optical, mcgrew2018atomic}.
These advancements required important breakthroughs, including the capability to cool and trap fermionic alkaline-earth atoms in spin-insensitive potentials\cite{takamoto2003spectroscopy, barber2006direct, ye2008quantum}; the development of ultra-coherent lasers\cite{kessler2012sub40mhzlinewidth, cole2013tenfold, matei2017mu} to fully exploit an ultra-narrow clock transition\cite{ludlow2015optical}; the detailed characterization of inter-atomic interactions\cite{scazza2014observation, cappellini2014direct}; and, more recently, the preparation of a quantum degenerate gas in a three-dimensional (3D) optical lattice\cite{campbell2017fermidegenerate, marti2018imaging, goban2018emergence}.
Nonetheless, all improvements in sensing capabilities to date have been based on single-particle control of internal atomic degrees of freedom.
Such strategies will eventually have diminishing returns due to practical difficulties in (i) suppressing decoherence from external (motional) degrees of freedom, and (ii) interrogating more particles without additional systematic errors from interactions\cite{martin2013quantum, ludlow2015optical, marti2018imaging}.

Pushing beyond the current independent-particle paradigm requires leveraging many-body quantum correlations.
Entangled states such as spin-squeezed states\cite{kitagawa1993squeezed, wineland1992spin, ma2011quantum} can enhance measurement sensitivity, i.e.~the uncertainty $\Delta\theta$ in the estimation of a parameter $\theta$, below the standard quantum limit $\Delta\theta\sim1/\sqrt{N}$ for $N$ uncorrelated particles\cite{itano1993quantum, degen2017quantum}.
The major challenge for progress in this direction is that generating entanglement requires interactions, which are generally undesirable because they degrade atomic coherence, thereby limiting clock performance\cite{swallows2011suppression, martin2013quantum, rey2014probing, ludlow2011coldcollisionshift, lemke2011wave, ludlow2015optical}.
In fact, the most precise and accurate optical lattice clocks were designed to operate with fermionic atoms in identical nuclear and electronic states in order to suppress collisional decoherence\cite{campbell2009probing, swallows2011suppression, campbell2017fermidegenerate}, as identical fermions cannot interact via the otherwise dominant ($s$-wave) collisions at ultracold temperatures.
However, an initially spin-polarized Fermi gas still exhibits interactions at later times due to spin-orbit coupling (SOC) that is induced by the laser that drives the clock transition (i.e.~the ``clock laser'')\cite{wall2016synthetic, kolkowitz2016spinorbitcoupled, livi2016synthetic, bromley2018dynamics}.
Specifically, the momentum kick imparted by this laser imprints a position-dependent phase that induces inhomogenous spin precession and generates spin dephasing, thereby making atoms distinguishable and vulnerable to collisions.
While a deep lattice can suppress SOC, it also intensifies the light scattering which currently limits the coherence time of the clock\cite{dorscher2018latticeinduced, goban2018emergence, hutson2019engineering}.

In this work, we describe a scheme that can potentially lead to metrological advances in state-of-the-art optical lattice clocks through direct use of quantum entanglement by harnessing the interplay between nominally undesirable collisions and SOC.
This scheme is made possible in the weak SOC regime by the formation of an interaction-energy gap that suppresses the SOC-induced population transfer from the exchange-symmetric Dicke manifold (spanned by spin-polarized and thus non-interacting states) to the remainder of Hilbert space.
In this regime, the non-equilibrium dynamical processes induced by SOC result in all-to-all collective spin interactions that naturally generate spin-squeezed states in a manner robust to realistic experimental imperfections.
To generate spin squeezing, our protocol only requires control over (i) the orientation of the clock laser and (ii) the optical lattice depth.
These controls are straightforward to incorporate into current 3D clock interrogation sequences without sacrificing atom numbers or coherence times.
Additionally, we show that by applying a modulated drive from the clock laser, one can further prepare states that saturate the Heisenberg limit $\Delta\theta \sim 1/N$ for phase sensitivity\cite{kitagawa1993squeezed, ma2011quantum, degen2017quantum}.

\section{From the Fermi-Hubbard model to one-axis and two-axis twisting}
\label{sec:theory}

@@

\begin{figure*}
\centering
\includegraphics[width=0.6\textwidth]{protocol_schematic.pdf}
\caption{{\bf Schematic of the setup for spin squeezing}.
({\bf a}) We consider $N$ fermionic atoms with two (pseudo-)spin components, represented by red and blue spheres, trapped in the ground band of an optical lattice (shown in 2D for the sake of presentation).
Atoms tunnel to neighboring sites at a rate $J$ and experience on-site interactions with strength $U$.
An external laser carrying a position dependent phase $e^{i\v k_{\t{L}}\cdot\v r}$ couples the spin states of the atoms.
({\bf b}) After a gauge transformation, different spin states exhibit different dispersion relations with a relative phase $\phi=k_{\text{L}} a$, where $a$ is the lattice spacing.
The external laser couples spin states with identical quasi-momenta $q$ in the gauge-transformed frame.
({\bf c}) If interactions are sufficiently weak, all motional degrees of freedom become frozen in momentum space, with atoms effectively pinned to fixed quasi-momentum modes $\v q$.
The dynamics on the frozen $\v q$-space lattice can then be mapped to a spin model in which collisional interactions correspond to a uniform, all-to-all ferromagnetic Heisenberg Hamiltonian with strength $U/L$, where $L$ is the total number of lattice sites.
({\bf d}) The spin dependence of the dispersion relation is captured by a mode-dependent axial field $B_q$ that generates inhomogeneous spin precession.
This axial field couples exchange-symmetric many-body Dicke states with total spin $S=N/2$ to spin-wave states with $S=N/2-1$.
The all-to-all interaction opens an energy gap $fU$ (with $f=N/L$ the filling fraction of spatial modes) between the Dicke states and the spin-wave states, which forbids population transfer between them in the weak-field limit.
({\bf e}) To generate spin squeezing via one-axis twisting, we initialize a product state with all spins polarized in $-z$ (i.e.~in $\ket{\dn}$), and apply a fast external laser pulse to rotate all spins into $x$.
We then let atoms freely evolve for a variable time $t$ (with a spin-echo pulse), after which the amount of spin squeezing can be determined experimentally from global spin measurements.
The spin-squeezed state can be used for a follow-up clock interrogation protocol (see Appendix \ref{sec:clock_interrogation}).
}
\label{fig:protocol_schematic}
\end{figure*}

An external laser with Rabi frequency $\Omega$ and wavenumber $k_{\t{L}}$ along the tunneling axis resonantly couples atoms' internal states through the Hamiltonian $\hat H_{\t{laser}}/\hbar = \sum_j \Omega e^{ik_{\t{L}}x_j} \hat c_{j,\up}^\dag \hat c_{j,\dn} + \t{h.c.}$, where $\hat c_{j\alpha}$ is a fermionic annihilation operator for an atom on site $j$ with internal state $\alpha\in\set{\up,\dn}$ and $x_j$ is the position of site $j$.
This laser imprints a position-dependent phase that equates to a momentum kick $k_{\t{L}}$ when an atom changes internal states by absorbing or emitting a photon, thereby generating spin-orbit coupling\cite{wall2016synthetic, livi2016synthetic}.
After absorbing the position dependence of the laser Hamiltonian into fermionic operators through the gauge transformation $\hat c_{j,\up}\to e^{ik_{\t{L}} x_j}\hat c_{j,\up}$, which makes $\hat H_{\t{laser}}$ spatially homogeneous, the atoms are well-described in the tight-binding limit by the Fermi-Hubbard Hamiltonian\cite{esslinger2010fermihubbard}
\begin{align}
  \hat H_{\t{FH}}^{(\phi)}/\hbar
  &= -J \sum_j \p{e^{i\phi} \hat c_{j,\up}^\dag \hat c_{j+1,\up}
  + \hat c_{j,\dn}^\dag \hat c_{j+1,\dn} + \text{h.c.}} \notag \\
  &\qquad + U \sum_j \hat n_{j,\up} \hat n_{j,\dn},
\label{eq:FermiHubbard}
\end{align}
where $J$ is the nearest-neighbor tunneling rate; the SOC angle $\phi\equiv k_{\t{L}}a$ determines the phase gained by spin-up atoms upon tunneling from site $j$ to site $j+1$ (in the gauge-transformed frame) with lattice spacing $a=x_{j+1}-x_j$; $U$ is the on-site interaction energy of two atoms; and $\hat n_{j\alpha}\equiv\hat c_{j\alpha}^\dag \hat c_{j\alpha}$ is a number operator.

@@

On its own, the collective Heisenberg term ($\sim\hat{\v S}\c\hat{\v S}$) in Eqn.~\eqref{eq:SpinModel} opens an energy gap $fU$, with $f\equiv N/L$ the filling fraction of spatial modes, between the collective Dicke states $\ket{S=N/2, M_S}$ and the remainder of Hilbert space\cite{rey2008manybody, martin2013quantum, norcia2018cavitymediated} with $S<N/2$.
Here $S$ and $M_S$ respectively label the eigenvalues of the collective spin operators $\hat{\v S}\c\hat{\v S}$ and $\hat{S}_\z$, with eigenvalues $S(S+1)$ for non-negative $S\in\set{N/2,N/2-1,\cdots}$ and $M_S\in\set{-S,-S+1,\cdots,S}$.
The axial field $B_q$ generally couples states within the Dicke manifold to states outside it.
In the weak SOC limit (i.e.~$B_q\ll fU$), however, the interaction energy gap suppresses population transfer between states with different total spin $S$ [see Fig.~\ref{fig:protocol_schematic}(d)].
In this regime, the virtual occupation of states outside the Dicke manifold can be accounted for perturbatively.
The symmetries of SOC as expressed in Eqn.~\eqref{eq:SpinModel} dictate that this treatment should yield powers of $\hat S_\z$ when projected onto the collective Dicke manifold at higher orders in perturbation theory.
At second order in perturbation theory (see Appendix \ref{sec:derivation_OAT}), we thus find that SOC effectively yields a one-axis twisting (OAT) model widely known to generate squeezing dynamics\cite{kitagawa1993squeezed, ma2011quantum}:
\begin{align}
  \hat H_{\t{eff}}/\hbar
  = -\frac{U}{L} \hat{\v S} \c \hat{\v S}
  - \overline B \hat S_\z + \chi \hat S_\z^2,
  &&
  \chi \equiv \f{\widetilde B^2}{(N-1)fU},
  \label{eq:H_eff}
\end{align}
where $\overline B\equiv\sum_q B_q/N$ is the mean and $\widetilde B^2\equiv\sum_q\p{B_q-\overline B}^2/N$ the variance of the axial field.
The effect of the $\sim\v S\c\v S$ term is to generate a relative phase between states with different total spin $S$ and thus has no effect on dynamics  restricted to a fixed $S$.
Note also that  the collective spin rotation from $\overline B \hat S_\z$  can be eliminated by going into a rotating frame or by using a spin echo.

The entire protocol for preparing a squeezed state via OAT, sketched out in Fig.~\ref{fig:protocol_schematic}(e), reduces to a standard Ramsey protocol with a spin echo: after initially preparing a spin-down (i.e.~$-\uv z$) polarized sample of ultracold atoms populating the lowest Bloch band of a lattice, a fast $\pi/2$ pulse is applied with the clock laser to rotate all spin vectors into $+\uv x$.
The atoms then freely evolve for a variable time $t$ (possibly with spin-echo $\pi$-pulses), after which the amount of metrologically useful spin squeezing, measured by the Ramsey squeezing parameter
\begin{align}
  \xi^2 \equiv
  \min_\theta \braket{\t{var}(\hat S^\perp_\theta)}
  \times N/\abs*{\braket{\hat{\v S}}}^2,
\end{align}
can be determined experimentally from global spin measurements.
Here $\braket{\hat{\v S}}$ is the mean collective spin vector and $\braket{\t{var}(\hat S^\perp_\theta)}$ is the variance of spin measurements along an axis orthogonal to $\braket{\hat{\v S}}$, parameterized by the angle $\theta\in[0,2\pi)$.
The above protocol concerns only the preparation of a spin-squeezed state, which would then be used as an input state for a follow-up clock interrogation protocol without SOC.
We discuss subtleties of interfacing the squeezing protocol with a subsequent clock interrogation protocol in Appendix \ref{sec:clock_interrogation}.

@@

The above scheme for OAT achieves optimal spin squeezing that scales as $\xi^2_{\t{opt}}\sim N^{-2/3}$ with minimal intervention, i.e.~a standard Ramsey protocol.
Further improvements upon this scheme can be made by introducing a time-dependent driving field that transforms the OAT Hamiltonian into a two-axis twisting (TAT) one.
While the OAT model initially generates squeezing faster than the TAT model, the squeezing generation rate of OAT (measured in dB per second) falls off with time, while the squeezing generation rate for TAT remains approximately constant until reaching Heisenberg-limited amount of spin squeezing with $\xi^2_{\t{opt}}\sim N^{-1}$\cite{kitagawa1993squeezed, supplementary_material}.

>>>>

\documentclass{nature}

\bibliographystyle{naturemag}

@@

@@

%%% text for affiliations and contributions
\newcommand{\JILA}{${}^{\ref{JILA}}$}
\newcommand{\CTQM}{${}^{\ref{CTQM}}$}
\newcommand{\both}{${}^{\ref{JILA},\ref{CTQM}}$}

\author{
P.~He\both\footnote{Authors P.H.~and M.A.P.~contributed equally to this work.},~
M.~A.~Perlin\both\footnotemark[\value{footnote}],~
S.~R.~Muleady\both,~
R.~J.~Lewis-Swan\both,~
R.~B.~Hutson\JILA,~
J.~Ye\JILA,~
and A.M.~Rey\both
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\maketitle

\begin{affiliations}
\item JILA, National Institute of Standards and Technology and Department of Physics, University of Colorado, Boulder, CO, 80309, USA
\label{JILA}
\item Center for Theory of Quantum Matter, University of Colorado, Boulder, CO, 80309, USA
\label{CTQM}
\end{affiliations}

\begin{abstract}
%One of the most important tasks in modern quantum science is to coherently control and entangle many-body systems, and to subsequently use these systems to realize powerful quantum technologies such as quantum-enhanced sensors.
%However, many-body entangled states are difficult to prepare and preserve since internal dynamics and external noise rapidly degrade any useful entanglement.
%Here, we introduce a protocol that counterintuitively exploits inhomogeneities, a typical source of dephasing in a many-body system, in combination with interactions to generate metrologically useful and robust many-body entangled states.
%Motivated by current limitations in state-of-the-art three-dimensional (3D) optical lattice clocks (OLCs) operating at quantum degeneracy, we use local interactions in a Hubbard model with spin-orbit coupling to achieve a spin-locking effect.
%In addition to prolonging inter-particle spin coherence, spin-locking transforms the dephasing effect of spin-orbit coupling into a collective spin-squeezing process that can be further enhanced by applying a modulated drive.
%Our protocol is fully compatible with state-of-the-art 3D OLC interrogation schemes and may be used to improve their sensitivity, which is currently limited by the intrinsic quantum noise of independent atoms.
%We demonstrate that even with realistic experimental imperfections, our protocol may generate $\sim10$--$14$ dB of spin squeezing in $\sim1$ second with $\sim10^2$--$10^4$ atoms.
%This capability exemplifies a new paradigm of using driven non-equilibrium systems to overcome current limitations in quantum metrology, allowing OLCs to enter a new regime of enhanced sensing with correlated quantum states.
One of the most important tasks in modern quantum science is to coherently control and entangle many-body systems, realizing powerful quantum technologies such as quantum-enhanced sensors.
However, many-body entangled states are difficult to prepare and preserve due to internal dynamics and external noise.
Here, we introduce a protocol that counterintuitively exploits interactions and inhomogeneities, typical sources of dephasing, to generate metrologically useful and robust many-body entangled states.
Our protocol is fully compatible with state-of-the-art three-dimensional (3D) optical lattice clock (OLC) interrogation schemes, and can break through limits to sensitivity set by the intrinsic quantum noise of independent atoms.
Even with realistic experimental imperfections, our protocol may generate $\sim10$--$14$ dB of spin squeezing in $\sim1$ second with $\sim10^2$--$10^4$ atoms.
This capability exemplifies a new paradigm of using non-equilibrium systems to overcome limitations in quantum metrology, entering a new regime of enhanced sensing with correlated quantum states.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

@@

Optical lattice clocks in particular have seen some of the most impressive developments in recent years, reaching record levels of precision ($\sim 3\times 10^{-19}$)\cite{campbell2017fermidegenerate, marti2018imaging} and accuracy ($\sim 1\times 10^{-18}$)\cite{bloom2014optical, mcgrew2018atomic}.
These advancements required important breakthroughs, including the capability to cool and trap fermionic alkaline-earth atoms in spin-insensitive potentials\cite{takamoto2003spectroscopy, barber2006direct, ye2008quantum}; the development of ultra-coherent lasers\cite{kessler2012sub40mhzlinewidth, cole2013tenfold, matei2017mu} to fully exploit an ultra-narrow clock transition\cite{ludlow2015optical}; the detailed characterization of inter-atomic interactions\cite{scazza2014observation, cappellini2014direct, zhang2014spectroscopic}; and, more recently, the preparation of a quantum degenerate gas in a three-dimensional (3D) optical lattice\cite{campbell2017fermidegenerate, marti2018imaging, goban2018emergence}.
Nonetheless, all improvements in sensing capabilities to date have been based on single-particle control of internal atomic degrees of freedom.
Such strategies will eventually have diminishing returns due to practical difficulties in (i) suppressing decoherence from external (motional) degrees of freedom, and (ii) interrogating more particles without additional systematic errors from interactions\cite{martin2013quantum, ludlow2015optical, marti2018imaging}.

Pushing beyond the current independent-particle paradigm requires leveraging many-body quantum correlations.
Entangled states such as spin-squeezed states\cite{kitagawa1993squeezed, wineland1992spin, ma2011quantum} can enhance measurement sensitivity, i.e.~the uncertainty $\Delta\theta$ in the estimation of a parameter $\theta$, below the standard quantum limit $\Delta\theta\sim1/\sqrt{N}$ for $N$ uncorrelated particles\cite{itano1993quantum, degen2017quantum}.
The major challenge for progress in this direction is that generating entanglement requires interactions, which are generally undesirable because they degrade atomic coherence, thereby limiting clock performance\cite{swallows2011suppression, martin2013quantum, rey2014probing, ludlow2011coldcollisionshift, lemke2011wave, ludlow2015optical}.
In fact, the most precise and accurate optical lattice clocks were designed to operate with fermionic atoms in identical nuclear and electronic states in order to suppress collisional decoherence\cite{campbell2009probing, swallows2011suppression, campbell2017fermidegenerate}, as identical fermions cannot interact via the otherwise dominant ($s$-wave) collisions at ultracold temperatures.
However, an initially spin-polarized Fermi gas still exhibits interactions at later times due to spin-orbit coupling (SOC) that is induced by the laser that drives the clock transition (i.e.~the ``clock laser'')\cite{wall2016synthetic, kolkowitz2016spinorbitcoupled, livi2016synthetic, bromley2018dynamics}.
Specifically, the momentum kick imparted by this laser imprints a position-dependent phase that induces inhomogeneous spin precession and generates spin dephasing, thereby making atoms distinguishable and vulnerable to collisions.
While a deep lattice can suppress SOC, it also intensifies the light scattering which currently limits the coherence time of the clock\cite{dorscher2018latticeinduced, goban2018emergence, hutson2019engineering}.

In this work, we describe a scheme that can lead to metrological advances in state-of-the-art optical lattice clocks through direct use of quantum entanglement by harnessing the interplay between nominally undesirable collisions and SOC.
This scheme is made possible in the weak SOC regime by the formation of an interaction-energy gap that suppresses the SOC-induced population transfer from the exchange-symmetric Dicke manifold (spanned by spin-polarized, and thus non-interacting states) to the remainder of Hilbert space.
Interactions thereby prolong inter-particle spin coherence through a spin-locking effect, which additionally transforms the dephasing effect of SOC into a collective spin squeezing process.
To generate spin squeezing, our protocol only requires control over (i) the orientation of the clock laser and (ii) the optical lattice depth.
These controls are straightforward to incorporate into current 3D clock interrogation sequences without sacrificing atom numbers or coherence times.
Additionally, we show that by applying a modulated drive from the clock laser, one can further prepare states that saturate the Heisenberg limit $\Delta\theta\sim1/N$ for phase sensitivity\cite{kitagawa1993squeezed, ma2011quantum, degen2017quantum}.

%The intriguing and non-trivial observation reported here is that weakly interacting fermions in a lattice with dynamics described by the Hubbard model --- which obey fermionic quantum statistics and interact solely via collisions --- can feature all-to-all collective spin interactions that naturally generate spin-squeezed states in a manner robust to realistic experimental imperfections.
%In the spirit of quantum simulation, our work illustrates that short-range interacting systems can be used to emulate  dynamical magnetic behaviors found in arrays of long-range interacting systems (such as arrays of trapped ions\cite{bohnet2016quantum} or cavity QED systems\cite{schleier-smith2010states, norcia2018cavitymediated} with interactions mediated by phonons or photons, respectively), thus enhancing their potential as a metrological resource.
%Although surprising, this result is in line with the recent understandings in quantum research that explore how the interplay between well understood simple phenomena can give rise to emergent new behaviors.

Despite an abundance of proof-of-principle experiments with entangled states\cite{degen2017quantum, pezze2018quantum}, so far only the remarkable example of LIGO\cite{aasi2013enhanced, abbott2016gw150914} has demonstrated a quantum advantage in a state-of-the-art quantum sensing or measurement system.
%By demonstrating that the precise control of interactions, which becomes possible in the new generation of 3D optical lattice systems with fully quantized motional degrees of freedom, can naturally give rise to metrologically useful correlated many-body fermionic states, our work opens a path to not only generate entanglement but also to use it for a quantum advantage in a world-class sensor.
The new generation of 3D optical lattice systems have fully quantized motional degrees of freedom\cite{campbell2017fermidegenerate}, allowing for precise control of collisional interactions.
We demonstrate how these interactions can naturally give rise to metrologically useful correlated many-body fermionic states, opening a path to not only generate entanglement, but also harness it to achieve a quantum advantage in a world-class sensor.
Such an advance will ultimately deliver gains to real-world applications including timekeeping, navigation, communication, and our understanding of the fundamental laws of nature.

% Despite numerous demonstration experiments on entangled states[CITE], with the exception of LIGO\cite{aasi2013enhanced, abbott2016gw150914} no quantum advantage has yet been realized in any state-of-the-art quantum sensing or measurement system.
% By demonstrating that the precise control of interactions, which becomes possible in the new generation of 3D optical lattice systems with quantized motional degrees of freedom in all three directions, can naturally give rise to metrologically useful correlated many-body fermionic states, our work opens a path to not only generate entanglement but also to apply it to world-class sensors which can ultimately deliver gains to real-world applications including timekeeping, navigation, communication as well as advances in our understanding of fundamental laws of nature.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

\subsection{Spin squeezing with the Fermi-Hubbard model.}

@@

\begin{spacing}{1}
\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{protocol_schematic.pdf}
\caption{{\bf Schematic of the setup for spin squeezing}.
({\bf a}) We consider $N$ fermionic atoms with two (pseudo-)spin components, represented by red and blue spheres, trapped in the ground band of an optical lattice (shown in 2D for the sake of presentation).
Atoms tunnel to neighboring sites at a rate $J$ and experience on-site interactions with strength $U$.
An external laser carrying a position dependent phase $e^{i\v k_{\t{L}}\cdot\v r}$ couples the spin states of the atoms.
({\bf b}) After a gauge transformation, different spin states exhibit different dispersion relations with a relative phase $\phi=k_{\text{L}} a$, where $a$ is the lattice spacing.
The external laser couples spin states with identical quasi-momenta $q$ in the gauge-transformed frame.
({\bf c}) If interactions are sufficiently weak, all motional degrees of freedom become frozen in momentum space, with atoms effectively pinned to fixed quasi-momentum modes $\v q$.
The dynamics on the frozen $\v q$-space lattice can then be mapped to a spin model in which collisional interactions correspond to a uniform, all-to-all ferromagnetic Heisenberg Hamiltonian with strength $U/L$, where $L$ is the total number of lattice sites.
({\bf d}) The spin dependence of the dispersion relation is captured by a mode-dependent axial field $B_q$ that generates inhomogeneous spin precession.
This axial field couples exchange-symmetric many-body Dicke states with total spin $S=N/2$ to spin-wave states with $S=N/2-1$.
The all-to-all interaction opens an energy gap $fU$ (with $f=N/L$ the filling fraction of spatial modes) between the Dicke states and the spin-wave states, which forbids population transfer between them in the weak-field limit.
({\bf e}) To generate spin squeezing via one-axis twisting, we initialize a product state with all spins polarized in $-z$ (i.e.~in $\ket{\dn}$), and apply a fast external laser pulse to rotate all spins into $x$.
We then let atoms freely evolve for a variable time $t$ (with a spin-echo pulse), after which the amount of spin squeezing can be determined experimentally from global spin measurements.
The spin-squeezed state can be used for a follow-up clock interrogation protocol (see Methods).
}
\label{fig:protocol_schematic}
\end{figure}
\end{spacing}

An external laser with Rabi frequency $\Omega$ and wavenumber $k_{\t{L}}$ along the tunneling axis resonantly couples atoms' internal states through the Hamiltonian $\hat H_{\t{laser}}/\hbar = \sum_j \Omega e^{ik_{\t{L}}x_j} \hat c_{j,\up}^\dag \hat c_{j,\dn} + \t{h.c.}$, where $\hat c_{j\alpha}$ is a fermionic annihilation operator for an atom on site $j$ with internal state $\alpha\in\set{\up,\dn}$ and $x_j$ is the position of site $j$.
This laser imprints a position-dependent phase that equates to a momentum kick $k_{\t{L}}$ when an atom changes internal states by absorbing or emitting a photon, thereby generating spin-orbit coupling\cite{wall2016synthetic, livi2016synthetic}.
After absorbing the position dependence of the laser Hamiltonian into fermionic operators through the gauge transformation $\hat c_{j,\up}\to e^{ik_{\t{L}} x_j}\hat c_{j,\up}$, which makes $\hat H_{\t{laser}}$ spatially homogeneous, the atoms are well-described in the tight-binding limit by the Fermi-Hubbard Hamiltonian\cite{esslinger2010fermihubbard}
\begin{align}
  \hat H_{\t{FH}}^{(\phi)}/\hbar
  = -J \sum_j \p{e^{i\phi} \hat c_{j,\up}^\dag \hat c_{j+1,\up}
  + \hat c_{j,\dn}^\dag \hat c_{j+1,\dn} + \text{h.c.}}
  + U \sum_j \hat n_{j,\up} \hat n_{j,\dn},
\label{eq:FermiHubbard}
\end{align}
where $J$ is the nearest-neighbor tunneling rate; the SOC angle $\phi\equiv k_{\t{L}}a$ determines the phase gained by spin-up atoms upon tunneling from site $j$ to site $j+1$ (in the gauge-transformed frame) with lattice spacing $a=x_{j+1}-x_j$; $U$ is the on-site interaction energy of two atoms; and $\hat n_{j\alpha}\equiv\hat c_{j\alpha}^\dag \hat c_{j\alpha}$ is a number operator.

@@

On its own, the collective Heisenberg term ($\sim\hat{\v S}\c\hat{\v S}$) in Eqn.~\eqref{eq:SpinModel} opens an energy gap $fU$, with $f\equiv N/L$ the filling fraction of spatial modes, between the collective Dicke states $\ket{S=N/2, M_S}$ and the remainder of Hilbert space\cite{rey2008manybody, martin2013quantum, norcia2018cavitymediated, smale2019observation} with $S<N/2$.
Here $S$ and $M_S$ respectively label the eigenvalues of the collective spin operators $\hat{\v S}\c\hat{\v S}$ and $\hat{S}_\z$, with eigenvalues $S(S+1)$ for non-negative $S\in\set{N/2,N/2-1,\cdots}$ and $M_S\in\set{-S,-S+1,\cdots,S}$.
The axial field $B_q$ generally couples states within the Dicke manifold to states outside it.
In the weak SOC limit (i.e.~$B_q\ll fU$), however, the interaction energy gap suppresses population transfer between states with different total spin $S$ [see Fig.~\ref{fig:protocol_schematic}(d)].
In this regime, the virtual occupation of states outside the Dicke manifold can be accounted for perturbatively.
The symmetries of SOC as expressed in Eqn.~\eqref{eq:SpinModel} dictate that this treatment should yield powers of $\hat S_\z$ when projected onto the collective Dicke manifold at higher orders in perturbation theory.
At second order in perturbation theory (see Methods), we thus find that SOC effectively yields a one-axis twisting (OAT) model widely known to generate squeezing dynamics\cite{kitagawa1993squeezed, ma2011quantum}:
\begin{align}
  \hat H_{\t{eff}}/\hbar
  = -\frac{U}{L} \hat{\v S} \c \hat{\v S}
  - \overline B \hat S_\z + \chi \hat S_\z^2,
  &&
  \chi \equiv \f{\widetilde B^2}{(N-1)fU},
  \label{eq:H_eff}
\end{align}
where $\overline B\equiv\sum_q B_q/N$ is the mean and $\widetilde B^2\equiv\sum_q\p{B_q-\overline B}^2/N$ the variance of the axial field.
The effect of the $\sim\v S\c\v S$ term is to generate a relative phase between states with different total spin $S$ and thus has no effect on dynamics  restricted to a fixed $S$.
Note also that  the collective spin rotation from $\overline B \hat S_\z$  can be eliminated by going into a rotating frame or by using a spin echo.

The entire protocol for preparing a squeezed state via OAT, sketched out in Fig.~\ref{fig:protocol_schematic}(e), reduces to a standard Ramsey protocol with a spin echo: after initially preparing a spin-down (i.e.~$-\uv z$) polarized sample of ultracold atoms populating the lowest Bloch band of a lattice, a fast $\pi/2$ pulse is applied with the clock laser to rotate all spin vectors into $+\uv x$.
The atoms then freely evolve for a variable time $t$ (possibly with spin-echo $\pi$-pulses), after which the amount of metrologically useful spin squeezing, measured by the Ramsey squeezing parameter
\begin{align}
  \xi^2 \equiv
  \min_\theta \braket{\t{var}(\hat S^\perp_\theta)}
  \times N/\abs*{\braket{\hat{\v S}}}^2,
\end{align}
can be determined experimentally from global spin measurements.
Here $\braket{\hat{\v S}}$ is the mean collective spin vector and $\braket{\t{var}(\hat S^\perp_\theta)}$ is the variance of spin measurements along an axis orthogonal to $\braket{\hat{\v S}}$, parameterized by the angle $\theta\in[0,2\pi)$.

The above protocol concerns only the preparation of a spin-squeezed state, which would then be used as an input state for a follow-up clock interrogation protocol without SOC.
While increasing the lattice depth to turn off SOC during clock interrogation is the simplest approach, this will limit the interrogation time due to light scattering (see discussion below).
Alternatively, it is possible to keep the same lattice depth used for the spin squeezing generation by adding a specific pulse sequence to suppress SOC.
See details in the Methods section.

\subsection{Model validity.}

@@

In realistic implementations, the Gaussian profile of the laser beams always introduces an additional effective harmonic potential that modifies the translational invariance assumed so far.
We present a detailed discussion of the role of the harmonic trap in the Supplementary Information, where we demonstrate that the addition of harmonic confinement barely modifies the achievable spin squeezing with currently accessible trapping frequencies.
We find that the existence of single-particle localized modes in the lattice with harmonic confinement\cite{rey2005ultracold, pupillo2006extended} helps to protect spin squeezing and shifts the optimal parameter window to $U/J\gtrsim 2$.

\subsection{Two-axis twisting.}

The above scheme for OAT achieves optimal spin squeezing that scales as $\xi^2_{\t{opt}}\sim N^{-2/3}$ with minimal intervention, i.e.~a standard Ramsey protocol.
Further improvements upon this scheme can be made by introducing a time-dependent driving field that transforms the OAT Hamiltonian into a two-axis twisting (TAT) one.
While the OAT model initially generates squeezing faster than the TAT model, the squeezing generation rate of OAT (measured in dB per second) falls off with time, while the squeezing generation rate for TAT remains approximately constant until reaching Heisenberg-limited amount of spin squeezing with $\xi^2_{\t{opt}}\sim N^{-1}$ (see the Supplementary Information)\cite{kitagawa1993squeezed}.
